Recalibrating Gravitational Wave Phenomenological Waveform Model

### 2023-06-23 src/tex/ms.tex
Finished section 3

Since we wish to optimize the model over the whole parameter space,
we need to compare multiple model generated waveforms with NR waveforms.
However, the mismatch is only defined for two input waveforms at a particular set of intrinsic parameters.
We therefore define the loss function as an average of training waveforms in two ways,
the simple average of mismatches and the normalized average of mismatches,  
\begin{align}
	\mathcal{L}_{\mathrm{ave}}&=\frac{1}{N}\sum_{i=1}^N\mathcal{M}_i \\
	\mathcal{L}_{\mathrm{nl}}&=\frac{1}{N}\sum_{i=1}^N\frac{\mathcal{M}_i}{\mathcal{M}_{i,\mathrm{ini}}},
\end{align}	
where $\mathcal{M}_i$ represents the mismatch of an individual training waveform,
$\mathcal{M}_{i,\mathrm{ini}}$ represents the initial mismatch of the individual
training waveform, and $N$ is the total number of training waveforms.

@@

To illustrate the effect of optimiziation on an individual waveform level, in Fig.
\ref{fig:0154} we plot the phase and amplitude of a particular waveform before and after optimization together with
the NR waveform taken directly from the SXS catalog. In the bottom panel one can see that,
compared to the original IMRPhenomD waveform, the optimized waveform has smaller residuals both in amplitude and
phase, particularly in the inspiral region where the amplitude displays a
$50\%$ reduction in error. For a fair comparison, we selected one of the testing
waveforms from the catalog presented in \citep{khan2016frequency}.\te{Which one?}

With the purpose of improving downstream tasks such as parameter estimation in
mind, the more relevant metric of improvement is the distribution of improvement
in mismatch over the entire training dataset. Figure~\ref{fig:q148} shows the
distribution of log-mismatch for the testing waveforms before and after the
optimization procedure. For generality, we use a constant PSD in our loss
function. One can see the distribution have been skewed toward lower mismatch,
where the peak of the distribution is shifted by approximately an order of
magnitude, and a median mismatch is reduced by 50\%. When using
$\mathcal{L}_{\mathrm{nl}}$, we observe a similar improvement with a 22.9\%
decrease in the median.

@@

Note that the IMRPhenomD model was initially constructed and fitted using the
{\zdethp} weighted mismatch. To ensure a fair comparison, we apply the same
methods using the {\zdethp} PSD exhibit superior improvement compared to the
unweighted mismatch. We find no significant qualitative and quantitative
difference between the two PSDs in the resulting distribution of mismatches.

To understand whether additional training data can further improve the
performance of the model, we include waveforms that are not presence in the
original IMRPhenomD calibration in our training dataset with parameters
tabulated in Tab. \ref{tab:q1248}. We specifically choose to use $q=2$ events
since we have abundant $q=2$ NR waveforms to validate the final result. The new
set of coefficients generated from this optimization process yields only
marginal improvements in the newly produced waveforms, as seen in Fig.
\ref{fig:q148_q1248_compare}. The high mismatch tail of the optimized
distribution remains comparable to the original distribution, indicating that
the original dataset is sufficient for this task. Similarly, utilizing the
{\zdethp} PSD to optimize the loss function with additional waveforms results in 
similar level of improvement.

To investigate the performance of recalibration over the source parameter space,
we plot the improvement of mismatch in log projected over the parameter space of
$q$ vs. $\chi_{\mathrm{PN}}$ in Fig. \ref{fig:ps_q148_qchi}. On the mass ratio
axis, we can see the waveforms with $q\le4$ show most consistent average
improvement. This is because that part of the parameter space is better covered
by the training waveform, as compared to testing waveforms from $q=4$ to $q=8$,
where some testing waveforms lie outside the source parameter space covered by
the training set of waveforms. On the spin axis, we can see the waveform with
$\chi_{\textrm{PN}}$ close to zero show the most consistent improvement, and we
start deviate from 0, there could be fluctuations in the average improvement. On
top of that, in the $q\le4$ region, there seems to be more consistent
improvement of waveforms with $\chi_{\textrm{PN}}<0$. This can be understood as
since the original waveform was developed for aligned spin system, the waveforms
with $\chi_{\textrm{PN}}$ is less well-fitted \kw{Varify this.}, hence there
could be more room for improvement. To show this, we plot the parameter space of
$\chi_1$ vs. $\chi_2$ in Fig.~\ref{fig:ps_q148}. Waveforms along the diagonal
axis, i.e. $\chi_1\approx\chi_2$, show good mismatch improvements as discussed
above. Meanwhile, the top-left ($\chi_1<0<\chi_2$) and bottom-right
($\chi_1>0>\chi_2$) regions respond to optimization differently. In the top-left
region, waveforms generally improve slightly with along optimization. However,
waveforms in the bottom-right region do not improve after optimization. Some
waveforms even turned worse after optimization.

 Given that the waveform model's ansatz may not be entirely compatible with NR,
and the optimization procedure is carried out over a distribution of waveforms
with varying source parameters, it is conceivable that different parts of the
source parameter space may not share the same set of optimal IMRPhenomD
parameters, mean there could be some trade-offs in accuracy between different
parts of the parameter space. If this is the cause of the lack of improvement in
the high mismatch tail of the distribution, segmenting the parameter space into
smaller subspaces should alleviate this problem. On the other hand, if the
ansatz lacks the correct parameterized form to capture the NR waveforms'
behavior as a function of the source parameters, the results will always be
biased, and we should not expect any improvement, even if we segment the
parameter space during training.

@@

We divided the parameter space into four regions to analyze the effect of the
recalibration procedure on each region separately
(Fig.~\ref{fig:ps_q148_quadrant}). The training waveforms used for fitting in
this scenario are listed in Tab.~\ref{tab:quadrants} and the loss functions 
are calculated using simple average of the mismatches. The top-left and
bottom-right regions have limited data for $q>4$, hence the results are only
valid up to $q\lesssim4$. From Fig.~\ref{fig:ps_q148_quadrant}, we observe that
equal-spin waveforms lying on the diagonal has great improvement. The
improvement above the diagonal is also substantial compared to recalibrate using
all waveforms at once, barring a few caused by some testing waveforms having
$q>4$. This can be seen clearer in Fig.~\ref{fig:all_quadrants}, which we see
that the optimized histogram shifts downward uniformly. On the other hand, the
improvement below the diagonal is negligible, indicating our method struggles to
further improve on top of the original IMRPhenomD waveform. Interestingly, we
can see in the top-right and bottom-left quadrant, the lower right corner shows
systematic worsening similar to the case of all waveforms. This hinted in
general, the IMRPhenomD waveform ansatz does not fit well with waveforms with
anti-aligned spin where the primary spin is positive.

\begin{figure}[t]
	\script{all_quadrants.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/all_quadrants.pdf}
	\caption{Distributions of mismatches in the top-left (Top) and bottom-right
	(Bottom) regions. We use a constant noise spectrum to calculate mismatch and
	$\mathcal{L}_{\mathrm{ave}}$ as the loss function. Waveforms in the
	top-left region generally improves while waveforms in the bottom-right
	region has very little improvement, as indicated by the median (dashed
	lines). \kw{Is the original referring to original IMRPhenomD or optimized
	using everything? Can we have a reference for optimizing using everything?}}
	\label{fig:all_quadrants}
\end{figure}

>>>>

Since we wish to optimize the model over the whole parameter space,
we need to compare multiple model generated waveforms with NR waveforms.
However, the mismatch is only defined for two input waveforms at a particular set of intrinsic parameters.
We therefore define the loss function as an average of training waveforms in two ways,
the simple average of mismatches and the normalized average of mismatches,  
\begin{align}\label{eqn:loss}
	\mathcal{L}_{\mathrm{ave}}&=\frac{1}{N}\sum_{i=1}^N\mathcal{M}_i \\
	\mathcal{L}_{\mathrm{nl}}&=\frac{1}{N}\sum_{i=1}^N\frac{\mathcal{M}_i}{\mathcal{M}_{i,\mathrm{ini}}},
\end{align}	
where $\mathcal{M}_i$ represents the mismatch of an individual training waveform,
$\mathcal{M}_{i,\mathrm{ini}}$ represents the initial mismatch of the individual
training waveform, and $N$ is the total number of training waveforms.

@@

To illustrate the effect of optimiziation on an individual waveform level, in Fig.
\ref{fig:0154} we plot the phase and amplitude of a particular waveform before and after optimization together with
the NR waveform taken directly from the SXS catalog. In the bottom panel one can see that,
compared to the original IMRPhenomD waveform, the optimized waveform has smaller residuals both in amplitude and
phase, particularly in the inspiral region where the amplitude displays a
$50\%$ reduction in error. For a fair comparison, we selected one of the testing
waveforms from the catalog presented in \citep{khan2016frequency}.\te{Which one? Please write the name from the table.}

With the purpose of improving downstream tasks such as parameter estimation in
mind, the more relevant metric of improvement is the distribution of improvement
in mismatch over the entire validation dataset. Figure~\ref{fig:q148} shows the
distribution of log-mismatches for the validation waveforms before and after the
optimization procedure. Here we show results using a constant PSD in our loss
function. One can see the distribution of the optimized waveform is skewed toward lower mismatches,
with the peak of the distribution being shifted by approximately an order of
magnitude, and a median mismatch being reduced by 50\% (see vertical dotted lines). When using
$\mathcal{L}_{\mathrm{nl}}$, we observe a less pronounced improvement with a 22.9\%
decrease in the median.

@@

Note that the IMRPhenomD model was initally tested using the
{\zdethp} weighted mismatch. We would therefore like to examine whether using the {\zdethp} PSD in our loss function
could lead to an improved to mismatch. Peforming the optimization again, we find no significant difference
between the results using the two PSDs in the distribution of mismatches.

To understand whether additional training data can further improve the
performance of the model, we include waveforms that are not present in the
original IMRPhenomD calibration in our training dataset; the parameters
can be found in Tab.~\ref{tab:q1248}. We specifically choose to use $q=2$ events
since we have abundant $q=2$ NR waveforms to validate the final result. The new
set of coefficients generated from this optimization process yields only
marginal improvements in the newly produced waveforms, as seen in Fig.
\ref{fig:q148_q1248_compare}. The high mismatch tail of the newly optimized
distribution remains comparable to the distribution from the first optimization, indicating that
the original dataset is sufficient for this task. Similarly, utilizing the
{\zdethp} PSD in the loss function together with additional waveforms results in a
similar level of improvement.

To investigate the performance of recalibration over the source parameter space,
we plot the improvement of the log-mismatch as a function of the parameter space
$q-\chi_{\mathrm{PN}}$ in Fig. \ref{fig:ps_q148_qchi}. Red points indicate that the waveform
is improved by the optimization procedure, while blue points indicate that the waveform mismatch
\textit{increases} during optimization.
We can see that waveforms with $q\le4$ show the most consistent average improvement.
This is likely due to the better coverage of training waveforms in that part of the parameter space (see Fig.~\ref{fig:intrin_space}).
On the spin axis, we can see the waveforms with
$\chi_{\textrm{PN}}\sim0$ show the most consistent improvement. 
When we move away from $\chi_{\textrm{PN}}\sim0$, the improvement fluctuates but exhibits an overall trend. 
This is particularly true in the $q\le4$ region, where we see a consistent
improvement of the waveform for $\chi_{\textrm{PN}}<0$. 
Since the original waveform was developed for aligned-spin systems, the waveforms
with opposite sign $\chi_{1,2}$ are likely less well-fit \kw{Verify this.}.
To show this, we plot the parameter space $\chi_1 - \chi_2$ in Fig.~\ref{fig:ps_q148}. 
Waveforms along the diagonal axis, $\chi_1\sim\chi_2$, show good mismatch improvements as discussed
above. Meanwhile, the top-left ($\chi_1<0<\chi_2$) and bottom-right
($\chi_1>0>\chi_2$) regions respond to the optimization differently. In the top-left
region, waveforms generally improve slightly after optimization. However,
waveforms in the bottom-right region do not improve after optimization.

Given that the waveform model's ansatz may not be entirely compatible with NR,
and the optimization procedure is carried out over a distribution of waveforms
with varying source parameters, it is conceivable that different parts of the
source parameter space may not share the same set of optimal IMRPhenomD
parameters.
This would mean that there are trade-offs in accuracy between different
parts of the parameter space. If this is the cause of the lack of improvement in
the high mismatch tail of the distribution, segmenting the parameter space into
smaller subspaces should alleviate this problem. On the other hand, if the
ansatz lacks the correct parameterized form to capture the NR waveforms'
behavior as a function of the source parameters, the results will always be
biased, and we should not expect any improvement, even if we segment the
parameter space during training.

@@

We divided the parameter space into four regions to analyze the effect of the
recalibration procedure on each region separately
(Fig.~\ref{fig:ps_q148_quadrant}). The training waveforms used for fitting in
this scenario are listed in Tab.~\ref{tab:quadrants} and the loss functions 
are calculated using simple average of the mismatches, $\mathcal{L}_{\mathrm{ave}}$. The top-left and
bottom-right regions have limited data for $q>4$, hence the results are only
valid up to $q\lesssim4$. From Fig.~\ref{fig:ps_q148_quadrant}, we observe that
equal-spin waveforms lying on the diagonal improve the most. The
improvement above the diagonal is also significantly more than when optimizing using
all waveforms at once. This can be seen clearly in the top panel of Fig.~\ref{fig:all_quadrants}, where we see
that the optimized histogram shifts to lower mismatches.\footnote{
	Note that the top panel of Fig.~\ref{fig:all_quadrants} specifically shows the
	distribution of mismatches for the top-left quadrant of Fig.~\ref{fig:ps_q148_quadrant}.
} On the other hand, the
improvement below the diagonal is negligible, indicating that our method struggles to
further improve on top of the original IMRPhenomD waveform. 
Interestingly, we can see in the bottom-right quadrant that the waveform does not show a significant change
to the overall mismatch (see bottom panel of Fig.~\ref{fig:all_quadrants}). 
This suggests that the IMRPhenomD waveform ansatz fits waveforms with
$\chi_1>0$ and $\chi_2<0$ less well.

\begin{figure}[t]
	\script{all_quadrants.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/all_quadrants.pdf}
	\caption{Distributions of mismatches in the top-left (Top) and bottom-right
	(Bottom) regions of Fig.~\ref{fig:ps_q148_quadrant}. We use a constant noise spectrum to calculate mismatch and
	$\mathcal{L}_{\mathrm{ave}}$ as the loss function. Waveforms in the
	top-left region generally improves while waveforms in the bottom-right
	region has very little improvement, as indicated by the median (dashed
	lines). \kw{Is the original referring to original IMRPhenomD or optimized
	using everything? Can we have a reference for optimizing using everything?}}
	\label{fig:all_quadrants}
\end{figure}
